---
title: 'Child Document'
tag: "Child document"
---

```{r}
values = params$breakdown
values_og = values

stage = names(values[1])
describer =  names(values[2])

values = values[values[,3] != '-',]
for(i in 3:5){
  values[[i]] = values[[i]] |> as.numeric()
}

# Create horizontal bar plot grouped by stage values.
fig1 = plotly::plot_ly(data = values,
                x = ~percent_stage,
                y = values[[1]],
                color = values[[2]],
                type = 'bar',
                orientation = 'h',
                customdata =  values[[2]],
                hovertemplate = '%{customdata}: %{x}% <extra></extra>',
                marker =list(hoverformat ="%{2}f")
                ) |>
  layout(barmode = 'stack',
         yaxis = list(title = stage),
         xaxis = list(title = 'Percentage of students'),
         hovermode = 'y unified',
         legend = list(title = list(text = describer) )
         )

data_cur = values[values[[1]] != 'All',]
fig2 = plotly::plot_ly(data = data_cur,
                       x = ~percent_describer,
                       y = data_cur[[2]],
                       color = data_cur[[1]],
                       type = 'bar',
                       customdata =  data_cur[[1]],
                       orientation = 'h',
                       hovertemplate = '%{customdata}: %{x}% <extra></extra>',
                       marker =list(hoverformat ="%{2}f")
                       ) |>
                         layout(barmode = 'stack',
                                yaxis = list(title = describer),
                                xaxis = list(title = 'Percentage of students'),
                                hovermode = 'y unified',
                                legend = list(title = list(text = describer) )
                                
                         )

show_table = values_og[,-ncol(values)]
names(show_table)[3:5] = c('Total students', paste0('Percentage of ', describer ),  paste0('Percentage of ', stage))
if(!do_SDC){
show_table[[4]] = show_table[[4]] |> round(2)
show_table[[5]] = show_table[[5]] |> round(2)  
}


```

## `r describer`

In this section we explore the composition of `r stage` relative to `r describer`.
Within `r describer` there are `r length(unique(values[[2]]))` different values, these are `r paste0(unique(values[[2]]), collapse = ', ')`.

Below we produce bar charts showing the composition of each level in `r stage` by `r describer` (*Split by `r describer`* tab) and the reverse of showing the distribution of `r stage` levels within each `r describer` category (*Split by `r stage`* tab). We also provide the table used to create these charts.

#####  {.tabset}

###### Split by `r describer`

`r fig1`

###### Split by `r stage`

`r fig2`

###### Table

```{r}
show_table |>  DT::datatable(extensions = 'Buttons',
                rownames = FALSE,
                options = list(scrollX = TRUE,
                               pageLength = min(nrow(values), 10),
                               dom = ifelse(nrow(values) <=10 ,'Bt','Btp'),
                               buttons = c('copy', 'csv', 'excel', 'pdf'),
                               ordering=F,
                               paging=ifelse(nrow(values) <=10 ,F,T)),
                filter = list(position = "top"))
```

##### {-}


***
