---
title: "Stage composition"
output: html_document
---

##### {-}
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(htmltools)
library(sf)
library(here)
library(flextable)
library(ggrepel)
library(LStools)
```

```{r, imported parameters}
#Inputs
# set.seed(33)
# KS1 = sample(c('3+','2A','2B', '2C-'), 10000, replace = T, prob = c(.21,.28,.26,.25))
# KS2 = sample(c('6','5+', '5-','4','3-'), 10000, replace = T, prob = c(0.03, 0.15, 0.22, 0.46, 0.13))
# GCSE = sample(c('9','8', '7', '6-'), 10000, replace = T, prob = c(0.03, 0.07, 0.10, 0.80))
# A_level =  sample(c('Maths + Further',  'Maths', 'No Maths'), 10000, replace = T, prob = c(0.02, 0.09, 0.89))
# FSM = sample(c('Yes', 'No'), 10000, replace = T, prob = c(.238, 1-.238))
# 
# raw = data.frame(FSM = factor(FSM),
#                  KS1 = factor(KS1, levels = c('3+','2A','2B', '2C-')),
#                  KS2 = factor(KS2, levels = c('6','5+', '5-','4','3-')),
#                  GCSE= factor(GCSE, levels = c('9','8', '7', '6-')), 
#                  A_level = factor(A_level, levels = c('Maths + Further',  'Maths', 'No Maths')))
# 
# 
# describer_columns = 1:3
# stage_column = 4
# report_name = 'GCSE composition'
# report_kind = 'interactive'
# reference_docx = NULL
# output_dir = getwd()
# do_SDC = TRUE

pathways = raw_to_pathway(raw)
if(is.numeric(stage_column)){
  stage_column = names(pathways)[stage_column]
}

if(is.numeric(describer_columns)){
  describer_columns = names(pathways)[describer_columns]
}

```

```{r, headline values without SDC, eval = !do_SDC}
total_students = sum(pathways$No_students)
missing_stage = sum(is.na(raw[[stage_column]]))

# Composition of all students.
stage = raw[[stage_column]]
stage = stage[!is.na(stage)]
stage_table = table(stage) |> data.frame()
stage_table$percent = (stage_table$Freq / sum(stage_table$Freq) * 100 ) |> round(2)
names(stage_table) = c('Stage value', 'Number of Students', 'Percent of Students')
below_minimum = which(stage_table$Freq == 1)
if(length(below_minimum)>0){
  stage_table[below_minimum, ]
}
## Composition by describers
breakdown = pbapply::pblapply(describer_columns, function(col_name){
  table_of_values = raw |>
    dplyr::count(dplyr::across(c(stage_column, col_name))) |>
    dplyr::group_by(dplyr::across(c(col_name))) |>
    dplyr::mutate(percent_describer = n / sum(n)*100) |>
    dplyr::group_by(dplyr::across(c(stage_column))) |>
    dplyr::mutate(percent_stage = n / sum(n)*100)
  table_of_values$describer = col_name
  
  describer_only_table = raw |>
    dplyr::count(dplyr::across(c(col_name))) |>
    dplyr::mutate(percent_describer = n / sum(n)*100) 
  describer_only_table$percent_stage = describer_only_table$percent_describer
  describer_only_table$describer = col_name
  describer_only_table[[stage_column]] = rep('All', nrow(describer_only_table))
  describer_only_table = describer_only_table[,match(names(table_of_values), names(describer_only_table))]
  
  rbind(describer_only_table, table_of_values)
})
names(breakdown) = describer_columns

```

```{r, headline values with SDC, eval = do_SDC}
SDC <- function(numbers, do_SDC = TRUE, round_to = 5, issue_level = 5, numeric_only = T){
  if(!do_SDC){
    return(numbers)
  }
  new_links = numbers
  below_minimum = which(new_links <= issue_level & new_links > 0)
  if(length(below_minimum) > 0){
     new_links[below_minimum] = ifelse(numeric_only, 0, paste0('Below ', issue_level))
     
     new_links[-below_minimum] = new_links[-below_minimum] |>
    as.numeric() |>
    plyr::round_any(round_to)

  return(new_links)
  }
 
  return(new_links |> plyr::round_any(round_to))
  
}

total_students = sum(pathways$No_students) |> SDC(do_SDC = do_SDC, issue_level = issue_level, numeric_only = F, round_to = round_to)
missing_stage = sum(is.na(raw[[stage_column]])) |> SDC(do_SDC = do_SDC, issue_level = issue_level, numeric_only = F, round_to = round_to) 


# Composition of all students.
stage = raw[[stage_column]]
stage = stage[!is.na(stage)]
stage_table = table(stage) |> data.frame()
text_stage_freq =  stage_table$Freq |>  SDC(do_SDC = do_SDC, issue_level = issue_level,numeric_only = F, round_to = round_to)
stage_table$Freq = stage_table$Freq |>  SDC(do_SDC = do_SDC, issue_level = issue_level, round_to = round_to)
stage_table$percent = (stage_table$Freq / sum(stage_table$Freq) * 100 ) |> round(0)
names(stage_table) = c('Stage value', 'Number of Students', 'Percent of Students')
below_minimum = grep('Below',text_stage_freq)
if(length(below_minimum)>0){
  stage_table[below_minimum, 2:3 ] = '-'
}
## Composition by describers
breakdown = pbapply::pblapply(describer_columns, function(col_name){
  table_of_values = raw |>
    dplyr::count(dplyr::across(c(stage_column, col_name)))
  text_n = table_of_values$n |>  SDC(do_SDC = do_SDC, issue_level = issue_level,numeric_only = F,  round_to = round_to)
  table_of_values$n = table_of_values$n |> SDC(do_SDC = do_SDC, issue_level = issue_level, round_to = round_to)
  table_of_values = table_of_values |>
    dplyr::group_by(dplyr::across(c(col_name))) |>
    dplyr::mutate(percent_describer = (n / sum(n)*100) |> round()) |>
    dplyr::group_by(dplyr::across(c(stage_column))) |>
    dplyr::mutate(percent_stage = (n / sum(n)*100)|> round(0)) |> as.data.frame()

  table_of_values$describer = col_name
  below_minimum = grep('Below',text_n)
  if(length(below_minimum)>0){
  table_of_values[below_minimum, 3:5 ] = '-'
  }
  
  describer_only_table = raw |>
    dplyr::count(dplyr::across(c(col_name)))
   text_n = describer_only_table$n |>  SDC(do_SDC = do_SDC, issue_level = issue_level,numeric_only = F, round_to = round_to)
  describer_only_table$n = describer_only_table$n |> SDC(do_SDC = do_SDC, issue_level = issue_level, round_to = round_to)
  
  describer_only_table = describer_only_table |>
    dplyr::mutate(percent_describer = (n / sum(n)*100)|> round(0)) 
  describer_only_table$percent_stage = describer_only_table$percent_describer
  describer_only_table$describer = col_name
  describer_only_table[[stage_column]] = rep('All', nrow(describer_only_table))
  describer_only_table = describer_only_table[,match(names(table_of_values), names(describer_only_table))]
  
    below_minimum = grep('Below',text_n)
  if(length(below_minimum)>0){
  describer_only_table[below_minimum, 3:5 ] = '-'
  }
  
  rbind(describer_only_table, table_of_values)
})
names(breakdown) = describer_columns

```

```{r, save values}
main_title = openxlsx::createStyle(fontSize = 14, textDecoration = "bold", halign = "left")
second_title = openxlsx::createStyle(fontSize = 10, textDecoration = "italic", halign = "left")
student_count = openxlsx::createStyle(numFmt = "#,##0")

# Create and save the excel file
titles = paste0('Number and percentage of students split by their scores/demographics in ', stage_column, ' and ', describer_columns)
wb <- openxlsx::createWorkbook()
for(i in 1:length(breakdown)){
  values_og = breakdown[[i]]
  show_table = values_og[,-ncol(values_og)]
  names(show_table)[3:5] = c('Total students', paste0('Percentage of ', describer_columns[i] ),  paste0('Percentage of ', stage_column))
  if(!do_SDC){
    show_table[[4]] = show_table[[4]] |> round(2)
    show_table[[5]] = show_table[[5]] |> round(2)  
  }
  
  sheet = describer_columns[i]
  openxlsx::addWorksheet(wb, sheet)
  openxlsx::writeData(wb, sheet, x = titles[i], startRow = 1, startCol = 1)
  openxlsx::writeDataTable(wb,
                           sheet = sheet,
                           x = show_table,
                           startRow = 4,
                           rowNames = FALSE,
                           withFilter = TRUE)
  openxlsx::addStyle(wb = wb, sheet = sheet, rows = 1, cols = 1, style = main_title)
  openxlsx::addStyle(wb = wb, sheet = sheet, rows = 2, cols = 1, style = second_title)
  openxlsx::addStyle(wb = wb, sheet = sheet, rows = 4:(nrow(show_table)+4), cols = 3, style = student_count)
  openxlsx::setColWidths(wb, sheet=sheet, cols = 2:5, widths = 'auto')
  
}


# Save workbook
openxlsx::saveWorkbook(wb,  file = paste0(output_dir,'/',stage_column, '_stage_composition.xlsx'), overwrite = TRUE)
```
This report analyses the composition of **`r stage_column`** and in particular we analyses the affect of **`r paste0(describer_columns, collapse = ', ')`** on the composition. 

Within the data we have **`r total_students`** students in total. Of these students **`r missing_stage`** (`r (missing_stage/ total_students *100) |> round(2)`%) have missing `r stage_column` information. 

Let's begin by looking at the composition of **`r stage_column`** across all students.

```{r all students table, eval = report_kind == 'interactive'}
stage_table |>  DT::datatable(#extensions = 'Buttons',
                rownames = FALSE,
                options = list(scrollX = TRUE,
                               pageLength = nrow(stage_table),
                               dom = 't',
                               #buttons = c('copy', 'csv', 'excel', 'pdf'),
                               paging=FALSE,
                               ordering=F))
```

In the following sections we explore the the interplay between `r stage_column` and `r describer_columns`.

***

<!-- Loop over all describer columns and add the information  -->

```{r breakdown by describers, results='asis'}
knit_child_with_params <- function(p) {
  params <- p
  knitr::knit_child(
    "child.rmd",
    envir = environment(),
    quiet = TRUE)
}

# This needs to be declared outside the function
input = list()
for(i in 1:length(describer_columns)){
  input[[i]] = list(breakdown = breakdown[[i]])
}
names(input) = describer_columns


# res = knit_child_with_params(input[[2]])
          
res <- lapply(input, knit_child_with_params)
# counter = 0
# res <- lapply(params, function(x){
#   counter <<- counter + 1
#   print(counter)
#  knit_child_with_params(x) 
# })


cat(unlist(res), sep = '\n')
```
