---
title: "Creating simple tables"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LStools)
```

*This vignette shows how to use the LStools package to create composition tables of variables from educational data, whether used for outputs or sanity checks within the SRS`*

To use the function we require some raw data, where each row corresponds to a unique student. We generate some data below where we have columns for Free school meal (FSM) status, maths grades at KS1, KS2 and GCSE and whether the student does a-level maths (or further maths).

```{r, echo = T}
set.seed(33)
KS1 = sample(c('3+','2A','2B', '2C-'), 10000, replace = T, prob = c(.21,.28,.26,.25))
KS2 = sample(c('6','5+', '5-','4','3-'), 10000, replace = T, prob = c(0.03, 0.15, 0.22, 0.46, 0.13))
GCSE = sample(c('9','8', '7', '6-'), 10000, replace = T, prob = c(0.03, 0.07, 0.10, 0.80))
A_level =  sample(c('Maths + Further',  'Maths', 'No Maths'), 10000, replace = T, prob = c(0.02, 0.09, 0.89))
FSM = sample(c('Yes', 'No'), 10000, replace = T, prob = c(.238, 1-.238))

raw = data.frame(FSM = factor(FSM),
                 KS1 = factor(KS1, levels = c('3+','2A','2B', '2C-')),
                 KS2 = factor(KS2, levels = c('6','5+', '5-','4','3-')),
                 GCSE= factor(GCSE, levels = c('9','8', '7', '6-')), 
                 A_level = factor(A_level, levels = c('Maths + Further',  'Maths', 'No Maths')))
```

`r DT::datatable(raw[1:10,], , rownames = FALSE, options = list(scrollX = TRUE,pageLength = nrow(10), dom = 'tp', paging=FALSE))`

Note that I have factored the data in such a way that the grades are ordered from highest to lowest for each column (where applicable), this isn't required but will affect the ordering of values in the outputted tables. 

Generally, we work with pathways which is a simplified form of the raw data consisting of only the unique combinations of values across all columns together with the number of students who follow each set of values.

```{r, echo = T}
pathways = LStools::raw_to_pathway(raw)
```

The first 10 rows of pathways are shown below

`r DT::datatable(pathways[1:10,], , rownames = FALSE, options = list(scrollX = TRUE,pageLength = nrow(10), dom = 'tp', paging=FALSE))`

### Composition of columns

Within LStools `pathway_summary_columns()` provides a useful sanity check is to look at the composition of each column within pathways, to check if for errors potentially created during the cleaning stage. The function creates a list of data frames for each column within pathways detailing the number and percentage of students split by the values in the column.

```{r}
composition_columns = pathways |> LStools::pathway_summary_columns()
composition_columns
```

***

### Composition tables

A common output from the SRS will be composition tables of stages of the maths pipeline. For example GCSE maths grades split by demographics of students. 

Within LStools `pathway_composition_tables()` can be used to create the composition tables for multiple stages across multiple demographic columns within pathways.

```{r, create composition table}
composition_by_FSM_KS1 = pathways |> pathway_composition_tables(describers = c('FSM','KS1'), stages = c('KS2', 'GCSE'))
composition_by_FSM_KS1
```

Prior to outputting from the SRS we need to apply statistical disclosure control which can be turned on via a toggle in `pathway_composition_tables()`.

```{r create composition table with SDC}
composition_by_FSM_KS1_SDC = pathways |> pathway_composition_tables(describers = c('FSM','KS1'), stages = c('KS2', 'GCSE'),do_SDC = TRUE) # By default SDC is turned off (FALSE).
composition_by_FSM_KS1_SDC
```

We can then save the composition table to excel ready to be checked for output. We will use `data_frame_to_workbook()` to allow each stage of the composition table to be unique sheets within an xlsx workbook each with an individual title and source information.

```{r save composition table, eval = FALSE}
data_frame_to_workbook(data = composition_by_FSM_KS1_SDC$KS2,
                            sheet = 'KS2',
                            title = 'Composition of KS2 maths grade by FSM status and KS1 maths grade',
                            source = 'Source: National Pupil Database') |>
  data_frame_to_workbook(data = composition_by_FSM_KS1_SDC$GCSE,
                              sheet = 'GCSE',
                              title = 'Composition of GCSE maths grade by FSM status and KS1 maths grade',
                              source = 'Source: National Pupil Database') |>
  openxlsx::saveWorkbook(file = 'composition.xlsx', overwrite = TRUE)
```


`pathway_composition_tables()` can also be used across multiple years of data allowing trends in composition to be explored. For example if our raw data had another column for the year the student was in year 11 (final year of secondary school) 

```{r, echo = T}
set.seed(33)
KS1 = sample(c('3+','2A','2B', '2C-'), 10000, replace = T, prob = c(.21,.28,.26,.25))
KS2 = sample(c('6','5+', '5-','4','3-'), 10000, replace = T, prob = c(0.03, 0.15, 0.22, 0.46, 0.13))
GCSE = sample(c('9','8', '7', '6-'), 10000, replace = T, prob = c(0.03, 0.07, 0.10, 0.80))
A_level =  sample(c('Maths + Further',  'Maths', 'No Maths'), 10000, replace = T, prob = c(0.02, 0.09, 0.89))
FSM = sample(c('Yes', 'No'), 10000, replace = T, prob = c(.238, 1-.238))
IDACI = sample(c('1st','2nd','3rd', '4th', '5th'), 10000, replace = T)
Year = sample(c('2022','2023'), 10000, replace = T, prob = c(.45, .55))
KS1[sample(1:10000,50)] = NA
raw_years = data.frame(Year = factor(Year),
                 FSM = factor(FSM),
                 IDACI = factor(IDACI),
                 KS1 = factor(KS1, levels = c('3+','2A','2B', '2C-')),
                 KS2 = factor(KS2, levels = c('6','5+', '5-','4','3-')),
                 GCSE= factor(GCSE, levels = c('9','8', '7', '6-')),
                 A_level = factor(A_level, levels = c('Maths + Further',  'Maths', 'No Maths'))
)

pathways_years = LStools::raw_to_pathway(raw_years)

```

we can use the variable `year_column` to specify the year in `pathway_composition_tables()` resulting in the following composition tables being created (note that this should only be used if the values (grades) remain the same between the years)

```{r, create composition table with years}
composition_by_FSM_IDACI_2022_23 = pathways_years |> pathway_composition_tables(describers = c('FSM','IDACI'), stages = c('KS2', 'GCSE'),year_column = 'Year')
composition_by_FSM_IDACI_2022_23$KS2 # Look only at KS2 composition
```

The table can then be used to look at trends over time of the composition of stages in the pipeline. For example in our fake dataset has the results of FSM students changed at KS2 between 2022 and 2023?

```{r dummy example}
required_data = composition_by_FSM_IDACI_2022_23$KS2 |> dplyr::filter(describer == 'FSM' & level == 'Yes')
required_data[,4:8] = (required_data[,4:8] / required_data[,9] *100) |> round(digits = 2)
required_data
```
***

### Composition of transitions

It might be of interest to explore the composition of students who 'flow' through a particular transition of the maths pipeline, such as those who achieve level 6 at KS2 who go on to achieve a grade 6 or worse at GCSE.

Within LStools `pathway_stage_transition_by_describer()` can be used to extract the demographics (or other quantiites) of students at all transitions through the mathematics pipeline. 

For example we can extract the FSM status of students through transitions through KS1, KS2 and GCSE with the following code

```{r extract FSM transitions}
transition_demographics = pathway_stage_transition_by_describer(pathways, describers = 'FSM', stages = c('KS1','KS2', 'GCSE'))
# transition_demographics contains 40 transitions in a list, show only two.
transition_demographics$`KS1: 3+ -> KS2: 6`
transition_demographics$`KS2: 5+ -> GCSE: 9`
```

By default the output is a list of transitions where each transition has a list for the composition of each demographic. This format is useful when wanting to concentrate on a single transition. To alter the output format we can change the value of `return_type` to 'compressed' which will instead return a data frame for each demographic showing the composition of all the transitions. This format is easier to output from the SRS.
```{r extract FSM transitions compressed output}
transition_demographics = pathway_stage_transition_by_describer(pathways, describers = c('FSM', 'A_level'), stages = c('KS1','KS2', 'GCSE'), return_type = 'compressed')
lapply(transition_demographics, head)
```

```{r save transition table, eval = FALSE}
data_frame_to_workbook(data = transition_demographics$FSM,
                            sheet = 'FSM',
                            title = 'Transitions between KS1, KS2 and GCSE split by students FSM status',
                            source = 'Source: National Pupil Database') |>
  data_frame_to_workbook(data = transition_demographics$A_level,
                              sheet = 'A-level',
                              title = 'Transitions between KS1, KS2 and GCSE split by students A-level mathematics choices',
                              source = 'Source: National Pupil Database') |>
  openxlsx::saveWorkbook(file = 'transition_info.xlsx', overwrite = TRUE)
```
