---
title: "Creating simple reports"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LStools)
```

*In this vignette we show how to use the LStools package to create simple reports from raw student data*.


### Composition of a stage 

The function `create_stage_composition_report()` can be used to simply create a report given some cleaned student information. Where the report concentrates on a single stage (KS1 maths, GCSE maths, etc) and looks at the composition of the students at that stage relative to other quantities (demographics, grades at other stages). 

To use the function we require some raw data, where each row corresponds to a unique student. We generate some data below where we have columns for Free school meal (FSM) status, maths grades at KS1, KS2 and GCSE and whether the student does a-level maths (or further maths).

```{r, echo = T}
raw = LStools::fake_data
```

`r DT::datatable(raw[1:10,], , rownames = FALSE, options = list(scrollX = TRUE,pageLength = nrow(10), dom = 'tp', paging=FALSE))`

Note that I have factored the data in such a way that the grades are ordered from highest to lowest for each column (where applicable), this is such that in the report the grades are shown in that order rather than alphabetical (Also useful for sankey diagrams, see vignette *Creating Sankey Diagrams*).

Suppose we want to look at the composition of GCSE maths grades relative to students achievement in mathematcs in KS1 and KS2, in addition to FSM status. 

We can create a report exploring this by running

```{r create report, eval = F}
create_stage_composition_report(raw = raw,
                         stage_column = 'GCSE',
                         describer_columns = c('KS1', 'KS2','FSM'),
                         output_dir = getwd(),
                         output_file = 'Composition_of_GCSE_by_KS1_KS2_FSM.html')

```

`create_stage_composition_report()` also allows for statistical disclosure control via the input `do_SDC` which can be fine tuned using the parameters `round_to` and `issue_level` to control the level of rounding and student counts to supress.

```{r create report SDC, eval = F}
create_stage_composition_report(raw = raw,
                         stage_column = 'GCSE',
                         describer_columns = c('KS1', 'KS2','FSM'),
                         do_SDC = TRUE,
                         round_to = 5,
                         issue_level = 5,
                         output_dir = getwd() 
                         output_file = 'Composition_of_GCSE_by_KS1_KS2_FSM_with_SDC.html')

```
